

# Methods


To see how mining coverage has impacted rivers we need two datasets. First
we need to know when mining first occured on the landscape. Second we need
to know how these sections of mined lands are connected to river networks and
how mining impacts might propagate downstream. The first dataset is part of the
Pericack et al., paper (2018) labeled `First Mining Year (GeoTIFF)` in the 
figshare data repository. The second dataset we will create here using `whitebox` and elevation data from the region, but fist we need to download the data. 

```{r setup, include=FALSE}
library(sf) #Amazingly simple tidy GIS R package
library(mapview) #Interactive mapping of sf objects
library(tidyverse) #Good 'ol tidyverse (dplyr, readr, more)
library(elevatr) #R access to mapzen (dynamic downloading of DEMs)
library(raster) # Name says it all, rasters in R
library(whitebox)
library(stars) # A bridge to the future of raster (spatiotemporal objects)
library(USAboundaries)
library(tmap)
knitr::opts_chunk$set(echo = TRUE)

# Run once!
#devtools::install_github("giswqs/whiteboxR")
```



```{r, include=F}

#Look for cumulative mining data
cume_file <- 'data/in/cume.tif'

#Check if the file already exists, if not run the commands below. 
cume_downloaded = !file.exists(cume_file)

```


## Data acquisition and organization

### Cumulative mining from Pericack et al., 2018

Here we are simply downloading the Pericack et al First Mined dataset from
figshare using a Rmarkdown trick where we ask "does this file exist locally?"
If it does, then this code chunk will be skipped. If it doesn't then the 
data will be downloaded to your local directory under the folder 'data'.

```{r,eval=cume_downloaded}

#Create a data repo if one doesn't exist.
dir.create('data')
dir.create('data/in')
dir.create('data/out')

#Download data locally. Link is from paper
download.file('https://ndownloader.figshare.com/files/11446991?private_link=e99954fc2876c6e96a7c',destfile=cume_file,method='libcurl',mode='wb')



```




### Download relevant elevation data

Our watershed analyses will be based on elevation data which we will 
get from elevatr which relies on mapzen elevation data. In the USA this 
data comes directly from USGS data. 

```{r, include=F}
raw_dem_file <- 'data/in/elev_raw.tif'
raw_dem_eval <- !file.exists(raw_dem_file)
```


```{r, eval=raw_dem_eval}
elev_raw <- get_elev_raster(cume,z=11)


writeRaster(elev_raw,raw_dem_file,overwrite=T)
```


### Reproject elevation data into same projection as cumulative mining tif. 


```{r}
proj_dem_file <- 'data/out/elev.tif'
reproject <- !file.exists(proj_dem_file)


```


```{r, eval=reproject}

elev <- projectRaster(elev_raw,cume)
#Save this elev data for whitebox
writeRaster(elev,proj_dem_file,overwrite=T)

```